---
title: "Analysis of albumin levels of patients from Tongji Hospital, Wuhan, China"
author: "Dariusz Brzezinski"
date: "`r Sys.Date()`"
output: 
  html_document: 
    number_sections: yes
    toc: yes
    toc_float: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F, cache = F)

library(formattable)
library(readxl)
library(summarytools)
library(ggplot2)
library(ggforce)
library(dplyr)
library(zoo)
library(ggpubr)
library(lme4)
library(plotly)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
```

```{r functions}
DPI=600
PLOT_SIZE=5
FONT_SIZE=18
PLOT_UNITS="in"
DEATH_COLOR <- "#EE6677"
SURVIVED_COLOR <- "#4477AA"
NORMAL_RANGE_COLOR <- "#000000"
OUTCOME_PALETTE <- c(DEATH_COLOR, SURVIVED_COLOR)

read_data <- function(file) {
  df <- read_excel(file)
  df$PATIENT_ID <- na.locf(df$PATIENT_ID)
  df <- na.locf(df)
  
  for (id in unique(df$PATIENT_ID)) {
    df[df$PATIENT_ID == id,"First_RE_DATE"] <- df[df$PATIENT_ID == id, "RE_DATE"][[1]][1]
  }
  
  df_all_samples <- df %>%
    mutate(gender = replace(gender, gender == 1, "Male"), 
           gender = replace(gender, gender == 2, "Female"),
           gender = factor(gender, levels = c("Male", "Female")),
           outcome = replace(outcome, outcome == 0, "Survived"), 
           outcome = replace(outcome, outcome == 1, "Died"),
           OutcomeNumber = replace(outcome, outcome == "Survived", 1), 
           OutcomeNumber = replace(OutcomeNumber, outcome == "Died", 0),
           OutcomeNumber = as.numeric(OutcomeNumber),
           outcome = factor(outcome, levels = c("Died", "Survived")),
           TimeSinceAdmission = as.numeric(difftime(RE_DATE, First_RE_DATE), units="days")) %>%
    rename(Outcome = outcome, Gender = gender, Age = age, TimeSinceAdmission=TimeSinceAdmission, 
           Albumin = albumin, Glucose = glucose)
  
  df_all_samples
}

get_last_samples <- function(df_all_samples) {
  df <- df_all_samples %>%
    group_by(PATIENT_ID) %>% 
    summarise(Albumin=last(Albumin), Outcome=last(Outcome), OutcomeNumber=last(OutcomeNumber),
              Glucose=last(Glucose), Gender=last(Gender), Age=last(Age), n=n(),
              TimeSinceAdmission=last(TimeSinceAdmission))
  
  df
}

get_first_samples <- function(df_all_samples) {
  df <- df_all_samples %>%
    group_by(PATIENT_ID) %>% 
    summarise(Albumin=first(Albumin), Outcome=first(Outcome), OutcomeNumber=first(OutcomeNumber),
              Glucose=first(Glucose), Gender=first(Gender), Age=first(Age), n=n(),
              TimeSinceAdmission=first(TimeSinceAdmission))
  
  df
}

save_plot <- function(p, name, width=PLOT_SIZE, height=PLOT_SIZE) {
  ggsave(paste0("plots/", name, ".svg"), p, width=width, height=height, units=PLOT_UNITS, dpi=DPI)
  ggsave(paste0("plots/", name, ".png"), p, width=width, height=height, units=PLOT_UNITS, dpi=DPI)
}
```

This notebook presents the plot scripts and statistical analyses presented in the paper "Molecular determinants of dexamethasone vascular transport in COVID-19 therapy" by Shabalin *et al.* Plots presented herein are interactive versions of those presented in the manuscript.

# Data characteristics

```{r read}
df_raw <- read_excel("data/time_series_375_prerpocess_en.xlsx")
df_all <- read_data("data/time_series_375_prerpocess_en.xlsx")
df_first <- get_first_samples(df_all)
df_last <- get_last_samples(df_all)


raw_patients <- length(unique(df_raw$PATIENT_ID[complete.cases(df_raw$PATIENT_ID)]))
albumin_patients <- nrow(df_last)
mortality_rate <- nrow(df_last[df_last$Outcome == "Died",])/nrow(df_last) 
mortality_rate_male <- nrow(df_last[df_last$Outcome == "Died" & df_last$Gender == "Male",])/nrow(df_last[df_last$Gender == "Male",]) 
mortality_rate_female <- nrow(df_last[df_last$Outcome == "Died" & df_last$Gender == "Female",])/nrow(df_last[df_last$Gender == "Female",]) 
```

The data for the study was taken from the supplementary materials of "An interpretable mortality prediction model for COVID-19 patients" Nat. Mach. Intell. 2, 283–288 (2020) by *Yan et al.* and describes COVID-19 patients admitted to Tongji Hospital, Wuhan, China between January 10 and February 18, 2020. The raw dataset contained information about `r raw_patients` patients, however only `r albumin_patients` patients that had their albumin levels measured at least once during their hospital stay, and those were of interest to the current study. The table below presents the basic statistics of clinical variables analyzed in this study. These and other statistics, unless stated otherwise, are calculated based on the last available blood sample of given patient.

```{r stats, results='asis'}
Tongji <- df_last %>% select(Outcome, Gender, Age, TimeSinceAdmission, Albumin, Glucose)
# knitr::kable(summary(df_formatted_for_stats))
dfSummary(Tongji, plain.ascii = FALSE, style = "grid", headings = FALSE,
          graph.magnif = 0.75, valid.col = FALSE, tmp.img.dir = "/tmp")
```

The overall mortality rate was `r round(mortality_rate*100,2)`%, whereas the mortality rate among male and female patients was `r round(mortality_rate_male*100,2)`% and `r round(mortality_rate_female*100,2)`%, respectively.

# Albumin levels

## Distributions and normality tests

```{r sina_plot_function}
draw_albumin_sina_plot <-function(df) {
  p <- ggplot(df, aes(x=Outcome, y=Albumin, color=Outcome)) + 
  geom_hline(aes(yintercept=35), colour=NORMAL_RANGE_COLOR, linetype="dashed") +
  geom_hline(aes(yintercept=55), colour=NORMAL_RANGE_COLOR, linetype="dashed") + 
    
  geom_sina(seed=21) + 
  stat_summary(fun.data = "mean_sdl",  fun.args = list(mult = 1), 
               geom = "pointrange", color = "black") +
  theme_classic() + 
  theme(legend.position = "none",
        strip.background = element_blank(),
        text = element_text(size=FONT_SIZE), 
        axis.text = element_text(colour = "black", size=FONT_SIZE), 
        axis.title.y = element_text(margin = margin(t = 0, r = 8, b = 0, l = 0))) + 
  xlab("") + 
  ylab ("Patient's albumin level (g/L)") +
  labs(color = "Outcome") + 
  ylim(13, 56) +
  scale_color_manual(values=OUTCOME_PALETTE)
  
  p
}

draw_qq_plot <- function(df, column, y_lab) {
  ggqqplot(df_last, x=column, color="Outcome", palette = c("#EE6677", "#4477AA")) + 
  labs(color = "Outcome", fill = "Outcome") + 
  theme(legend.position = c(0.88, 0.18), 
        axis.text = element_text(colour = "black", size=18), 
        text = element_text(size=18)) + xlab("Theoretical quantiles") + ylab("Sample quantiles")
}
```

The sina plot below presents the distribution of albumin levels among the two outcome groups of patients (Died, Survived), with the mean and standard deviation overlaid. The horizontal dashed lines represent the normal range for albumin (35-55 g/L).

```{r albumin_sina_plot, cache=F}
albumin_sina_plot <- draw_albumin_sina_plot(df_last)
save_plot(albumin_sina_plot, "albumin_sina_plot")
ggplotly(albumin_sina_plot)
# ggplotly(albumin_sina_plot, tooltip = c("Gender", "Age", "Outcome", "TimeSinceAdmission", "Albumin", "Glucose"))
```

To verify whether the albumin levels in both outcome groups are normally distributed, we first plotted Q-Q plots, as presented below. The shaded area represents the 95% confidence intervals.

```{r qq_plot}
albumin_qq_plot <- draw_qq_plot(df_last, "Albumin")
save_plot(albumin_qq_plot, "albumin_qq_plot")
ggplotly(albumin_qq_plot)
```

Upon a positive verification of the Q-Q plots, the normality of the the albumin distributions was additionally confirmed by two Shapiro-Wilk tests at $\alpha = 0.05$. The null hypothesis for the test is that the distribution is normal. Therefore, if we *cannot* reject the null hypotheses, we will assume the the distributions are normal.

```{r shapiro}
format_test <- function(test_result, data_name, use_color=TRUE) {
  df <- data.frame(value = unlist(test_result))
  df["data.name", "value"] = data_name
  
  if (use_color){
    if (as.numeric(df["p.value", "value"]) < 0.05) {
      color = "#71CA97"
    } else {
      color = "#FA614B"
    }
    
    formattable(df, list(area(row = "p.value") ~ color_tile(color, color)))
  } else {
    formattable(df)
  }
}
```

### {.tabset -}

#### Shapiro-Wilk test: Died{-}
```{r}
format_test(shapiro.test(df_last[df_last$Outcome=="Died",]$Albumin), "Outcome == Died", use_color=FALSE)
```

#### Shapiro-Wilk test: Survived{-}
```{r}
format_test(shapiro.test(df_last[df_last$Outcome=="Survived",]$Albumin), "Outcome == Survived", use_color=FALSE)
```

### {-}

Since the p-values of the Shapiro-Wilk tests are above 0.05, we cannot reject the null hypothesis that the albumin levels are normally distributed. Therefore, we assume the null hypothesis is true and **the albumin levels for both outcome groups are normally distributed**. Having verified the normality of the distributions, we can perform a two-tailed Welch's t-test to check if the differences in albumin levels are statistically significant.

```{r t_test}
format_test(t.test(df_last[df_last$Outcome=="Died",]$Albumin, 
                                             df_last[df_last$Outcome=="Survived",]$Albumin), "Outcome == Died vs Outcome == Survived")
```


With a p-value < 0.001 we can reject the null hypothesis that the means are equal, and state that **the differences in mean albumin levels between the patients that died and survived are statistically significant**. The mean albumin level for those that sirvived was `r mean(df_last[df_last$Outcome=="Survived",]$Albumin)` g/L, whereas for those that died it was `r mean(df_last[df_last$Outcome=="Died",]$Albumin)` g/L.

## Gender

The difference between albumin levels was also evaluated for gender. It can be noticed that the albumin distribution shapes and means are practically identical to those of the overall patient cohort. Nevertheless, the proportions of points within those that died and survived for both genders are different. This relates to the earlier mentioned differences in mortality rates: `r round(mortality_rate_male*100,2)`% and `r round(mortality_rate_female*100,2)`% for males and females respectively)

```{r gender}
albumin_gender_plot <- albumin_sina_plot + facet_grid(cols = vars(Gender))
save_plot(albumin_gender_plot + theme(strip.text.x = element_blank()), "albumin_gender_plot")
ggplotly(albumin_gender_plot)
```

The differences in mean albumin levels within patients of the same gender were statistically significant according to a two-tailed Welch's t-test at $\alpha = 0.05$ (p < 0.001).

### {.tabset -}

#### Shapiro-Wilk test: Male Died{-}
```{r}
format_test(shapiro.test(df_last[df_last$Outcome=="Died" & df_last$Gender=="Male",]$Albumin), "Male: Outcome == Died", use_color=FALSE)
```

#### Shapiro-Wilk test: Male Survived{-}
```{r}
format_test(shapiro.test(df_last[df_last$Outcome=="Survived" & df_last$Gender=="Male",]$Albumin), "Male: Outcome == Survived", use_color=FALSE)
```

#### Welch's t-test: Male {-}
```{r}
format_test(t.test(df_last[df_last$Outcome=="Died" & df_last$Gender=="Male",]$Albumin, df_last[df_last$Outcome=="Survived" & df_last$Gender=="Male",]$Albumin), "Male: Outcome == Died vs Outcome == Survived")
```

### {.tabset -}

#### Shapiro-Wilk test: Female Died{-}
```{r}
format_test(shapiro.test(df_last[df_last$Outcome=="Died" & df_last$Gender=="Female",]$Albumin), "Female: Outcome == Died", use_color=FALSE)
```

#### Shapiro-Wilk test: Female Survived{-}
```{r}
format_test(shapiro.test(df_last[df_last$Outcome=="Survived" & df_last$Gender=="Female",]$Albumin), "Female: Outcome == Survived", use_color=FALSE)
```

#### Welch's t-test: Female {-}
```{r}
format_test(t.test(df_last[df_last$Outcome=="Died" & df_last$Gender=="Female",]$Albumin, df_last[df_last$Outcome=="Survived" & df_last$Gender=="Female",]$Albumin), "Female: Outcome == Died vs Outcome == Survived")
```

## Time series and trends

```{r}
draw_albumin_time_series_plot <- function(df_all, aDeath, bDeath, aSurvived, bSurvived) {
  p <- ggplot(df_all, aes(x=TimeSinceAdmission, y=Albumin, color=Outcome)) + 
    geom_hline(aes(yintercept=35), colour=NORMAL_RANGE_COLOR, linetype="dashed") +
    geom_hline(aes(yintercept=55), colour=NORMAL_RANGE_COLOR, linetype="dashed") +
    geom_line(mapping=aes(group=PATIENT_ID), alpha=0.35) +
    geom_segment(aes(x=0, xend=22, y=bDeath + aDeath,
                     yend = bDeath + aDeath*22),
                 color=DEATH_COLOR, size=1.25) +
    geom_segment(aes(x=0, xend=22, y=bSurvived + aSurvived,
                     yend = bSurvived + aSurvived*22),
                 color=SURVIVED_COLOR, size=1.25) +
    theme_classic() + 
    theme(legend.position = c(0.88, 0.18),
          text = element_text(size=FONT_SIZE), 
          axis.text = element_text(colour = "black", size=FONT_SIZE), 
          axis.text.y = element_blank()) +
    xlab("Days since admission") + 
    ylab ("") +
    labs(color = "Outcome") + 
    ylim(13, 56) +
    scale_color_manual(values=OUTCOME_PALETTE)
  

  p
}
```

```{r}
d <- rbind(df_first, df_last)

fits <- lmList(Albumin ~ TimeSinceAdmission | PATIENT_ID, data=d)
coefs <- coef(fits)
coefs <- cbind(coefs, df_last)

aDeath <- median(coefs[coefs$Outcome == "Died", "TimeSinceAdmission"], na.rm=T)
bDeath <- median(coefs[coefs$Outcome == "Died", "(Intercept)"], na.rm=T)
aSurvived <- median(coefs[coefs$Outcome == "Survived", "TimeSinceAdmission"], na.rm=T)
bSurvived <- median(coefs[coefs$Outcome == "Survived", "(Intercept)"], na.rm=T)
```


Apart from studying the patients' final albumin samples, we have also analyzed how albumin levels changed over time. The line plot below presents the albumin level of each patient over time. The median linear trends between the first and last albumin levels recorded were:

$$ albumin_{Died} = `r round(aDeath, 2)`x + `r bDeath`$$
$$ albumin_{Survived} = `r round(aSurvived, 2)`x + `r bSurvived`$$


```{r}
albumin_time_series_plot <- draw_albumin_time_series_plot(df_all, aDeath, bDeath, aSurvived, bSurvived)
save_plot(albumin_time_series_plot + theme(legend.position = "none"), "albumin_time_series_plot")
albumin_time_series_plot + ylab("Patient's albumin level (g/L)") + theme(axis.text.y = element_text(colour = "black", size=FONT_SIZE))
```

Calculating Pearson's correlation coefficient, we can state that the correlation between albumin levels and days since hospital admission is statisticaly significant only for those patients that died. 

### {.tabset -}

#### Pearson's Correlation Coefficient: Died{-}
```{r}
format_test(cor.test(d[d$Outcome == "Died", ]$TimeSinceAdmission, d[d$Outcome == "Died", ]$Albumin, method = c("pearson")), 
            "Outcome == Died")
```

#### Pearson's Correlation Coefficient: Survived{-}
```{r}
format_test(cor.test(d[d$Outcome == "Survived", ]$TimeSinceAdmission, d[d$Outcome == "Survived", ]$Albumin, method = c("pearson")),
            "Outcome == Survived")
```


## Albumin vs glucose levels

Another variable that was taken into account was the pateint's glucose level. The scatter plot below presents the relation between albumin levels (y-axis) and glucose levels (x-axis), with color denoting the outcome groups of patients (red: Died, blue: Survived).The horizontal dashed lines represent the normal range for albumin (35-55 g/L), whereas the vertical dashed lines show the glucose fasting normal range (4.0-5.5 mmol/L) and the random plasma test diabetes threshold (11.1 mmol/L).

```{r}
draw_albumin_glucose_plot <-function(df) {
  p <- ggplot(df, aes(x=Glucose, y=Albumin, color=Outcome, fill=Outcome, 
                      text = paste("Gender:", Gender, "<br>","Age:", Age, "<br>",
                                   "Albumin:", Albumin, "g/L<br>","Glucose:", Glucose, "mmol/L"))) + 
  geom_hline(aes(yintercept=35), colour=NORMAL_RANGE_COLOR, linetype="dashed") +
  geom_hline(aes(yintercept=55), colour=NORMAL_RANGE_COLOR, linetype="dashed") +
  geom_vline(aes(xintercept=4.0), colour=NORMAL_RANGE_COLOR, linetype="dashed") + 
  geom_vline(aes(xintercept=5.5), colour=NORMAL_RANGE_COLOR, linetype="dashed") + 
  geom_vline(aes(xintercept=11.1), colour=NORMAL_RANGE_COLOR, linetype="dashed") + 
  geom_point() + 
  theme_classic() + 
  theme(legend.position = c(0.85, 0.88), 
        text = element_text(size=FONT_SIZE), 
        axis.text.y = element_blank(),
        axis.text = element_text(colour = "black", size=FONT_SIZE)) + 
  xlab("Patient's glucose level (mmol/L)") + 
  ylab("") +
  labs(color = "Outcome", fill = "Outcome") + 
  ylim(13, 56) +
  scale_color_manual(values=OUTCOME_PALETTE) +
    scale_fill_manual(values=OUTCOME_PALETTE)
  
  p
}

albumin_glucose_plot <- draw_albumin_glucose_plot(df_last)
save_plot(albumin_glucose_plot + theme(legend.position="none"), "albumin_glucose_plot", width=PLOT_SIZE*2, height = PLOT_SIZE)
ggplotly(albumin_glucose_plot + ylab("Patient's albumin level (g/L)") + theme(axis.text.y = element_text(colour = "black", size=FONT_SIZE)), tooltip = c("text"))
```

It can be noticed that both albumin and glucose levels can be associated with COVID-19 outcome. The relation will be analyzed as part of the logistic regression analysis.

## Albumin vs age

Similarly to glucose levels, the scatter plot below presents the relation between albumin levels (y-axis) and age (x-axis), with color denoting the outcome groups of patients (red: Died, blue: Survived).

```{r}
draw_albumin_age_plot <-function(df) {
  p <- ggplot(df, aes(x=Age, y=Albumin, color=Outcome, 
                      text = paste("Gender:", Gender, "<br>","Age:", Age, "<br>",
                                   "Albumin:", Albumin, "g/L<br>","Glucose:", Glucose, "mmol/L"))) + 
  geom_hline(aes(yintercept=35), colour=NORMAL_RANGE_COLOR, linetype="dashed") +
  geom_hline(aes(yintercept=55), colour=NORMAL_RANGE_COLOR, linetype="dashed") +
  geom_point() + 
  theme_classic() + 
  theme(legend.position = "none", 
        text = element_text(size=FONT_SIZE), 
        axis.text = element_text(colour = "black", size=FONT_SIZE),
        axis.title.y = element_text(margin = margin(t = 0, r = 8, b = 0, l = 0))) + 
  xlab("Patient's age") + 
  ylab ("Patient's albumin level (g/L)") +
  labs(color = "Outcome") + 
  ylim(13, 56) +
  scale_color_manual(values=OUTCOME_PALETTE)
  
  p
}

albumin_age_plot <- draw_albumin_age_plot(df_last)
save_plot(albumin_age_plot + theme(legend.position="none"), "albumin_age_plot")
ggplotly(albumin_age_plot, tooltip = c("text"))
```

Once again, it can be noticed that age can be associated with albumin levels and COVID-19 outcome. This relation will also be analyzed as part of the logistic regression analysis.

# Regression analysis

To verify the statistical significance of associations between different variables analyzed in this study (albumin levels, glucose levels, gender, age) and the outcome (Died, Survived). We will first created unadjusted models to see the relations between each single variable and the outcome. After that we will create and adjusted model where all the variables are taken into account as potential confounders. Finally, we will verify whether apart from confounding there is any effect modification between albumin levels and other variables.

```{r}
run_logistic_regression <- function(df, formula) {
  logistic.fit<-glm(formula, family="binomial", data=df)
  # print(summary(logistic.fit))
  # print(exp(logistic.fit$coefficients))
  # print(exp(confint(logistic.fit)))
  tab_model(logistic.fit, p.style = "numeric_stars") %>% 
    return() %$% 
    knitr %>% 
    knitr::asis_output()
}
```

## Unadjusted models

Below are the results of logistic regression for each variable. All the unadjusted models were found to be statistically significant at $\alpha = 0.05$, with p < 0.001.

### Albumin{-}

```{r}
run_logistic_regression(df_last, OutcomeNumber~Albumin)
```

### Gender{-}

```{r}
run_logistic_regression(df_last, OutcomeNumber~Gender)
```

### Age{-}

```{r}
run_logistic_regression(df_last, OutcomeNumber~Age)
```

### Glucose{-}

```{r}
run_logistic_regression(df_last, OutcomeNumber~Glucose)
```

## Adjusted model

The adjusted model has shown that albumin levels are statistically significantly associated with COVID-19 outcome (p < 0.001), even when confounding factors are taken into account.

```{r}
run_logistic_regression(df_last, OutcomeNumber~Albumin + Glucose + Age + Gender)
```

## Testing for effect modification

Test for interactions between albumin levels and other variables, did not show any significant effect modification; all p > 0.20.

### Albumin and gender{-}
```{r}
run_logistic_regression(df_last, OutcomeNumber~Albumin*Gender)
```

### Albumin and age{-}
```{r}
run_logistic_regression(df_last, OutcomeNumber~Albumin*Age)
```

### Albumin and glucose{-}
```{r}
run_logistic_regression(df_last, OutcomeNumber~Albumin*Glucose)
```

## Regression based on gender

Finally, for the only categorical confounding variable we have plotted (gender) a logistic regression plot with separate lines for males and females. It can be noted that gender makes a different only for high albumin levels. In other words, low albumin levels are an equally strong predictor of death from COVID-19.

```{r}
gender_albumin_regression <- ggplot(df_last, aes(x=Albumin, y=OutcomeNumber, color=Gender)) + 
  geom_point() + 
  geom_vline(aes(xintercept=35), colour=NORMAL_RANGE_COLOR, linetype="dashed") +
  geom_vline(aes(xintercept=55), colour=NORMAL_RANGE_COLOR, linetype="dashed") +
  theme_classic() + 
  theme(legend.position = c(0.15, 0.13),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        text = element_text(size=FONT_SIZE), 
        axis.text = element_text(colour = "black", size=FONT_SIZE), 
        axis.text.y = element_blank()) + 
  ylab("Probability of death") + 
  xlab ("") +
  xlim(13, 56) +
  labs(color = "Gender") + 
  scale_color_manual(values=c("#44AA99", "#AA4499")) +
  stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE) +
  coord_flip()
save_plot(gender_albumin_regression, "gender_albumin_regression")
ggplotly(gender_albumin_regression  + xlab("Patient's albumin level (g/L)") + theme(axis.text.y = element_text(colour = "black", size=FONT_SIZE)))
```

# Citing

If you find this analysis useful, please cite: Ivan G. Shabalin1, Mateusz P. Czub, Karolina A. Majorek, Dariusz Brzezinski, Marek Grabowski, David R. Cooper, Mateusz Panasiuk, Maksymilian Chruszcz, Wladek Minor, “Molecular determinants of dexamethasone vascular transport in COVID-19 therapy”, *in review*.
